Stochastic blockmodeling of relational event dynamics

### 2012-06-01 docs/inference.tex
Small notation fixes.
Minor edits and clean up.

We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
\begin{align*}
p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
\end{align*}
where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.
Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.

@@

\subsection{Hyperparameter settings}
In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2000} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.  This aims for similarly sized groups.}
We set $\alpha_{\sigma}=3$ and $\beta_{\sigma}=1$ so that in the presence of little data we encourage shrinkage towards the upper level parameters $\mu$.  
As with other models, we note that the predictive accuracy of the model can depend on the hyperparameters.

>>>>

We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
\begin{align*}
p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{U}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
\end{align*}
where $\mathcal{U}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.
Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.

@@

\subsection{Hyperparameter settings}
In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2000} with 5 extra clusters drawn from the prior.%\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.  This aims for similarly sized groups.}
We set $\alpha_{\sigma}=5$ and $\beta_{\sigma}=1$ so that in the presence of little data we encourage shrinkage towards the upper level parameters $\mu$.  
As with other models, we note that the predictive accuracy of the model can depend on the hyperparameters.

### 2012-06-01 docs/model.tex
Minor edits and clean up.

Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001,Kemp2006} for static networks which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

The proposed family of models generalizes several important special cases.
For example,  using only the intercept statistic $s_0(t,i,j,\mathcal{A}_t) = 1$ is analogous to the stochastic block model for static networks.
Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
As these intensities do not change under this specification, the likelihood simplifies to
$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$

Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
\begin{align}
\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
\label{eqn:multllk}
\end{align}
The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1973}, though here the possible choices are all the dyads in $\mathcal{R}$.

@@

\subsection{Related work}
Stochastic blockmodels have been extended to longitudinal network data
involving a sequence of networks occurring at discrete times
\cite{Ishiguro2010, Rodriguez2011}, while we focus on methods for
analyzing a stream of relational events. 
The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}. % \cite{AalenOddO.2008} 
Relational event models such as these require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions.  
We instead use latent variables to allow for heterogeneity in intensities across the set of possible events.

>>>>

Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001} for static networks which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

The proposed family of models generalizes several important special cases.
For example,  using only the intercept statistic $s_0(t,i,j,\mathcal{A}_t) = 1$ is analogous to the stochastic block model for static networks.
Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
As these intensities do not change under this specification, the likelihood simplifies to
\begin{align}
\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m}
\prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}
\end{align}
Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
\begin{align}
\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
\label{eqn:multllk}
\end{align}
The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1973}, though here the possible choices are all the dyads in $\mathcal{R}$.

@@

\subsection{Related work}
Stochastic blockmodels have been extended to longitudinal network data
involving a sequence of networks occurring at discrete times
\cite{Ishiguro2010, Rodriguez2011}, while we focus on methods for
analyzing a stream of relational events. 
The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2011,Vu2011,Vu2011a}. % \cite{AalenOddO.2008} Opsahl2011,Stadtfeld2010
Relational event models such as these require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions.  
We instead use latent variables to allow for heterogeneity in intensities across the set of possible events.
